\documentclass{article}
\usepackage[final]{aaj2026}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}

\title{AIOps-Enabled Agent2Agent Communication Framework for Automated ITIL Service Management}

\author{
  Luis Rubio, MSc \\
  DevSecOps/Cloud Solution Architect \\
  Ericsson Spain \\
  \texttt{luis.rubio@ericsson.com} \\
}

\begin{document}

\maketitle

\begin{abstract}
Artificial Intelligence for IT Operations (AIOps) represents a paradigm shift in Digital Product and Service Management (DPSM), leveraging machine learning, natural language processing, and automated reasoning to enhance operational efficiency and reduce human intervention in routine IT tasks. AIOps platforms integrate diverse data sources, apply advanced analytics to identify patterns and anomalies, and provide intelligent insights for proactive incident management, performance optimization, and capacity planning. The convergence of AI technologies with traditional IT operations enables organizations to achieve higher service availability, faster incident resolution, and improved resource utilization through predictive analytics and automated decision-making processes. The evolution toward Agentic AIOps introduces autonomous intelligent agents that can independently execute complex IT operations tasks, make contextual decisions, and collaborate through standardized communication protocols to deliver comprehensive service management automation.

The integration of AIOps capabilities with ITIL v5 practices offers significant potential for automating core service management processes, including incident management, problem management, change management, and service request fulfillment. Through intelligent agent-based architectures, Agentic AIOps can automate ticket classification and routing, perform root cause analysis, orchestrate remediation workflows, and maintain comprehensive service catalogs. This automation reduces manual effort, minimizes human error, and enables IT teams to focus on strategic initiatives while ensuring consistent adherence to ITIL best practices through standardized, AI-driven processes that continuously learn and adapt to organizational patterns. The agentic approach enables autonomous decision-making and inter-agent collaboration, creating self-managing IT ecosystems that can respond to complex scenarios without human intervention.
\end{abstract}

\section{Introduction}

This research explores the application of Agentic AIOps through the implementation of an Agent2Agent (A2A) communication framework designed to demonstrate automated Digital Product and Service Management (DPSM). The study investigates how autonomous intelligent agents, developed using LangGraph for workflow orchestration, Model Context Protocol (MCP) for standardized tool integration, and Agent2Agent (A2A) protocol for inter-agent communication, can be applied to automate ITIL v5 practices, specifically focusing on Monitoring and Event Management.

\textbf{LangGraph} is a framework for building stateful, multi-actor applications with Large Language Models (LLMs), enabling the creation of complex agent workflows through graph-based state management. It provides sophisticated control flow mechanisms, conditional routing, and persistent state management that allows agents to maintain context across multiple interactions and coordinate complex multi-step operations.

\textbf{Model Context Protocol (MCP)} is an open standard that enables secure connections between host applications and external data sources and tools. MCP provides a standardized interface for AI agents to interact with various systems, databases, and APIs while maintaining security boundaries and enabling seamless integration across different platforms and services.

\textbf{Agent2Agent (A2A) Protocol} facilitates standardized communication between autonomous agents through well-defined message formats and interaction patterns. A2A enables agents to discover each other's capabilities, exchange structured information, and coordinate complex workflows across distributed systems while maintaining loose coupling and scalability.

\section{Related Work}

The intersection of artificial intelligence and IT operations has generated significant research interest, particularly in the areas of autonomous agents, intelligent monitoring systems, and automated service management. This section reviews key contributions in Agentic AIOps and related domains.

\textbf{Autonomous Agent Systems in IT Operations}: Recent advances in autonomous agent architectures have demonstrated significant potential for IT operations automation. Chen et al. (2024) present a comprehensive framework for multi-agent systems in cloud infrastructure management, highlighting the importance of agent coordination and communication protocols for distributed IT environments \cite{chen2024}. Their work establishes foundational principles for agent-based automation that align with our A2A protocol implementation.

\textbf{AIOps and Machine Learning for Digital Product and Service Management (DPSM)}: The application of machine learning techniques to IT operations has been extensively studied. Kumar and Singh (2023) investigate the use of deep learning models for anomaly detection in large-scale distributed systems, demonstrating improved accuracy in incident prediction and root cause analysis \cite{kumar2023}. Similarly, Wang et al. (2024) explore reinforcement learning approaches for automated incident response, showing promising results in reducing mean time to resolution \cite{wang2024}.

\textbf{ITIL Framework Automation}: Several studies have examined the automation of ITIL processes through AI technologies. Rodriguez and Martinez (2023) propose an intelligent ticketing system that leverages natural language processing for automated ticket classification and routing, achieving 85\% accuracy in production environments \cite{rodriguez2023}. Their findings support the viability of AI-driven ITIL process automation, complementing our agent-based approach.

\textbf{Multi-Agent Communication Protocols}: The development of standardized communication protocols for agent systems has been a focus of recent research. Thompson et al. (2024) introduce a novel protocol for agent-to-agent communication in distributed computing environments, emphasizing security and scalability considerations \cite{thompson2024}. Their protocol design principles inform our implementation of the A2A communication framework.

\textbf{Event Management and Monitoring Automation}: Automated event management has received considerable attention in the literature. Li and Zhang (2023) present an AI-powered monitoring system that combines machine learning with rule-based approaches for intelligent event correlation and prioritization \cite{li2023}. Their hybrid approach demonstrates the effectiveness of combining multiple AI techniques for complex monitoring scenarios.

\textbf{Human-AI Collaboration in IT Operations}: The balance between automation and human oversight in IT operations has been explored by several researchers. Johnson et al. (2024) investigate human-AI collaboration patterns in DevOps environments, identifying key factors for successful integration of AI agents with human operators \cite{johnson2024}. Their findings emphasize the importance of transparency and explainability in agent decision-making processes.

While existing research has made significant contributions to individual aspects of AIOps and agent systems, there remains a gap in comprehensive frameworks that integrate specialized agents with standardized communication protocols for ITIL v5 practice automation. Our work addresses this gap by demonstrating a practical implementation of Agentic AIOps specifically designed for Monitoring and Event Management practices.

\section{Case Study: ITIL v5 Monitoring and Event Management Practice}

\subsection{Practice Overview}

Monitoring and Event Management is a critical ITIL v5 practice that constantly observes services within the organization and records all associated events. These events represent changes of state that impact the product's service delivery capabilities. The practice encompasses continuous surveillance of IT infrastructure, applications, and services to detect deviations from normal operating conditions. Monitoring and Event Management is particularly valuable in identifying information security events and facilitating rapid response with appropriate remediation solutions.

\subsection{Use Case Context: TechFlow Solutions}

\textit{Note: TechFlow Solutions is an invented company used solely to illustrate the use case and demonstrate the practical application of the proposed framework.}

To demonstrate the practical application of Agentic AIOps in ITIL v5 Monitoring and Event Management, this research examines the operational challenges faced by TechFlow Solutions, a mid-sized technology services company operating a cloud-native infrastructure. TechFlow Solutions maintains a production Kubernetes cluster hosting multiple microservices applications that serve over 10,000 daily active users across their e-commerce and customer relationship management platforms.

\textbf{Infrastructure Monitoring Scope}: The company's Kubernetes cluster consists of 15 worker nodes running approximately 200 pods across various namespaces including production applications, identity management services, and data processing pipelines. The cluster generates continuous streams of events related to pod lifecycle changes, resource utilization metrics, service health checks, and security alerts that require constant monitoring and analysis.

\textbf{Event Management System}: TechFlow Solutions utilizes a centralized service desk web portal for incident and service request management, implementing ITIL v5 best practices for ticket lifecycle management. The portal handles approximately 150-200 tickets per day, ranging from infrastructure alerts and application performance issues to user access requests and change management workflows. Operations staff must manually correlate Kubernetes cluster events with service desk tickets, often leading to delayed incident response and inefficient resource allocation.

\textbf{Operational Challenges}: The current manual approach requires dedicated personnel to continuously monitor cluster dashboards, analyze event logs, and create appropriate service desk tickets when issues are detected. This process is prone to human error, suffers from alert fatigue during high-volume periods, and lacks the contextual correlation needed to identify complex multi-system incidents before they impact end users.

\subsection{Human Complexity Challenges}

Traditional human-driven monitoring and event management faces significant complexity challenges that limit operational effectiveness:

\textbf{Volume and Velocity Overwhelm}: Modern IT environments generate thousands of events per minute across distributed systems, creating information overload that exceeds human processing capabilities. Operations teams struggle to distinguish critical alerts from routine notifications, leading to alert fatigue and missed incidents.

\textbf{Contextual Correlation Complexity}: Events often occur across multiple systems simultaneously, requiring deep technical knowledge to correlate seemingly unrelated alerts into coherent incident patterns. Human operators must maintain extensive mental models of system interdependencies, which becomes increasingly difficult as infrastructure complexity grows.

\textbf{24/7 Operational Demands}: Continuous monitoring requires round-the-clock staffing with consistent expertise levels, creating resource constraints and potential coverage gaps during shift transitions or staff unavailability.

\textbf{Response Time Pressures}: Critical events demand immediate analysis and response, but human decision-making under pressure can lead to suboptimal choices, delayed responses, or escalation of minor issues into major incidents.

\subsection{AI-Driven Improvements}

AI technologies address these human limitations through intelligent automation and augmentation:

\textbf{Intelligent Event Filtering and Prioritization}: Machine learning algorithms analyze historical event patterns to automatically classify and prioritize alerts based on business impact, reducing noise and focusing human attention on truly critical issues.

\textbf{Automated Correlation and Root Cause Analysis}: AI systems can process multiple event streams simultaneously, identifying complex correlation patterns across distributed systems that would be impossible for humans to detect manually, enabling faster root cause identification.

\textbf{Predictive Analytics}: Advanced algorithms analyze trends and patterns to predict potential failures before they occur, shifting from reactive to proactive monitoring approaches that prevent service disruptions.

\textbf{Continuous Learning and Adaptation}: AI systems continuously learn from resolved incidents, improving their accuracy in event classification, correlation, and response recommendations over time without human intervention.

\textbf{Autonomous Response Orchestration}: For well-defined scenarios, AI can automatically execute remediation workflows, reducing mean time to resolution and ensuring consistent response procedures regardless of time or staffing levels.

\section{Proposed Design}

The proposed Agentic AIOps framework adopts a specialized agent architecture where each agent is designed to perform a single, well-defined task within the IT operations domain. This design philosophy follows the principle of separation of concerns, enabling each agent to develop deep expertise in its specific operational area while maintaining loose coupling with other system components.

\textbf{Specialized Agent Approach}: Each agent in the framework is purpose-built for a specific ITIL practice or operational function. The Kubernetes monitoring agent focuses exclusively on infrastructure oversight and cluster management, while the ticketing agent specializes in incident and service request management. This specialization allows agents to optimize their knowledge base, tool integration, and response patterns for their designated domain, resulting in more accurate and efficient task execution.

\textbf{MCP Integration for External Systems}: Agents leverage the Model Context Protocol (MCP) to establish secure, standardized connections with external systems and data sources. MCP provides a unified interface that abstracts the complexity of different APIs and protocols, enabling agents to seamlessly integrate with diverse infrastructure components, monitoring tools, and enterprise systems without requiring custom integration code for each external service.

\textbf{A2A Protocol Orchestration}: The Agent2Agent (A2A) protocol serves as the coordination mechanism that enables specialized agents to work together as a cohesive system. Through A2A, agents can discover each other's capabilities, exchange structured messages, and coordinate complex workflows that span multiple operational domains. This orchestration layer maintains agent autonomy while enabling sophisticated multi-agent collaboration patterns.

The framework consists of three primary components that collectively demonstrate these Agentic AIOps principles applied to ITIL v5 practices:

\begin{enumerate}
\item \textbf{Kubernetes A2A Agent}: Implements infrastructure monitoring and management capabilities through MCP protocol integration for cluster resource oversight
\item \textbf{Ticketing A2A Agent}: Provides automated incident and service request management aligned with ITIL ticketing workflows
\item \textbf{Autonomous Agent Orchestrator}: Functions as an intelligent orchestration layer that autonomously plans and executes multi-agent workflows, enabling complex cross-domain operations without human intervention
\end{enumerate}

The implementation leverages LLM for natural language understanding, autonomous workflow planning, and demonstrates standardized inter-agent communication patterns through A2A protocol integration, enabling seamless coordination between specialized service management agents with autonomous decision-making capabilities.

\subsection{System Architecture}

The proposed Agentic AIOps framework implements a distributed multi-agent architecture with autonomous orchestration capabilities:

\textbf{Autonomous Agent Orchestrator (Central Intelligence)}:
\begin{itemize}
\item Implements LLM-based autonomous workflow planning for complex multi-step operations
\item Analyzes agent capabilities through exposed agent cards and skill definitions
\item Creates dynamic execution plans that span multiple specialized agents
\item Evaluates conditional logic based on intermediate results
\item Maintains conversation history and context across workflow steps
\item Provides intelligent context passing between agent invocations
\item Supports both single-agent and multi-agent coordination patterns
\item Handles agent discovery and capability resolution via A2A protocol
\end{itemize}

\textbf{Kubernetes Monitoring Agent}:
\begin{itemize}
\item Integrates with Kubernetes MCP server for real-time cluster monitoring
\item Exposes skills for resource monitoring, pod management, and cluster status
\item Implements streaming capabilities for continuous monitoring
\item Provides RESTful A2A endpoints for inter-agent communication
\end{itemize}

\textbf{Ticketing Service Agent}:
\begin{itemize}
\item Connects to ticketing backend via HTTP API
\item Implements CRUD operations for ticket management
\item Supports ticket creation, querying, and status tracking
\item Provides ITIL-compliant incident management workflows
\end{itemize}

\subsection{Communication Flow}

\textbf{Request Processing Pipeline}:
\begin{enumerate}
\item User submits natural language request to Autonomous Orchestrator
\item Orchestrator uses LLM to analyze request and create multi-step workflow plan
\item Workflow plan identifies required agents, actions, and conditional logic
\item For each workflow step:
  \begin{enumerate}
  \item Orchestrator evaluates conditions based on previous results
  \item If conditions are met, formats request as A2A protocol message
  \item Target agent processes request using LangChain agent executor
  \item Result is captured and stored as context for subsequent steps
  \end{enumerate}
\item Orchestrator synthesizes final response from all workflow results
\item Comprehensive summary is presented to user
\end{enumerate}

\subsection{Agent Implementation Details}

\textbf{Kubernetes Agent Architecture}:
\begin{itemize}
\item \textbf{MCP Integration}: Connects to Kubernetes MCP server on port 8080
\item \textbf{Tool Loading}: Dynamically loads Kubernetes management tools via MCP
\item \textbf{Session Management}: Maintains persistent MCP client sessions
\item \textbf{Error Handling}: Implements robust error handling for MCP connection failures
\end{itemize}

\textbf{Ticketing Agent Architecture}:
\begin{itemize}
\item \textbf{HTTP API Integration}: Direct REST API calls to ticketing backend on port 5000
\item \textbf{Tool Registration}: Implements create\_ticket, get\_all\_tickets, and query\_tickets tools
\item \textbf{LangChain Integration}: Uses Gemini LLM with specialized ticketing prompt
\item \textbf{ITIL Compliance}: Supports standard ITIL incident management processes
\end{itemize}

\textbf{Agent Host Architecture}:
\begin{itemize}
\item \textbf{Multi-Agent Coordination}: Manages connections to multiple specialized agents
\item \textbf{Request Classification}: Implements keyword-based routing with fallback logic
\item \textbf{A2A Client Management}: Maintains A2A clients for each registered agent
\item \textbf{Async Processing}: Supports asynchronous request processing and response handling
\end{itemize}

\section{Solution Implementation}

The Agentic AIOps framework is implemented in Python and hosted in a GitHub repository at \url{https://github.com/LuisRubio-zz/unie-aiops}. The implementation consists of specialized Python modules that demonstrate the A2A protocol integration with MCP-enabled agents for automated Digital Product and Service Management (DPSM).

\subsection{Environment Setup and Dependencies}

The framework requires Python 3.8+ and utilizes the \texttt{uv} package manager for dependency management. Key dependencies include LangChain for agent orchestration, FastAPI for A2A server implementation, and the MCP client library for Kubernetes integration.

\subsection{Agent Deployment}

The distributed nature of the Agentic AIOps framework requires each component to be executed in separate terminals to enable concurrent operation and inter-agent communication. This multi-terminal approach allows the ticketing backend, MCP server, specialized agents, and orchestration layer to run simultaneously as independent processes while maintaining A2A protocol connectivity.

\section{Results and Validation}

The framework demonstrates successful autonomous multi-agent orchestration with intelligent workflow planning and execution. The autonomous orchestrator successfully demonstrates multi-step workflow planning and execution, conditional logic based on intermediate results, context passing between agents, intelligent decision-making without human intervention, and graceful handling of both simple and complex requests.

\textbf{Autonomous Multi-Step Workflow Execution}: When presented with the request "List namespaces and create ticket if pods in error in Failed state", the system automatically decomposed the task into two steps: (1) querying the Kubernetes agent for pod status across all namespaces, and (2) conditionally creating a ticket with error details if issues were found. The system successfully identified 2 pods in Failed state and automatically created Ticket \#42 to track remediation.

\textbf{Conditional Workflow Execution}: For the request "Check cluster health and report issues", the orchestrator planned a 3-step workflow with conditional execution. When the first step determined all pods were running normally, the subsequent steps were automatically skipped, demonstrating intelligent decision-making based on intermediate results.

\textbf{Single-Agent Request Processing}: Simple requests such as "create a ticket for server maintenance" were efficiently routed to the appropriate agent without unnecessary orchestration overhead, showing the system's ability to adapt its complexity to the task at hand.

\section{Discussion}

Despite the notable progress of Agentic AIOps technologies in ITIL v5 Monitoring and Event Management, significant challenges remain, especially in dynamic IT environments characterized by high-volume event streams and complex infrastructure dependencies. Agents often struggle with interpreting evolving operational goals, adapting to changing infrastructure contexts, and processing multi-modal inputs from diverse monitoring sources.

A key tension in Agentic AIOps systems for monitoring and event management is balancing autonomous incident response with human oversight to ensure accurate, transparent, and efficient operations. This becomes increasingly important as agents take on roles that influence service availability and business-critical system stability.

To address these limitations in ITIL v5 monitoring contexts, future Agentic AIOps frameworks must integrate real-time adaptability to changing infrastructure patterns, context-aware reasoning for multi-system event correlation, and privacy-preserving mechanisms for sensitive operational data. As IT operations evolve toward more autonomous systems, the emphasis is shifting from pure automation toward human-centric AI collaboration, where optimal partnership between operations teams and monitoring agents is essential for maintaining service quality and operational excellence.

\section{Future Work}

The Agentic AIOps framework presented in this research provides a foundation for advanced Digital Product and Service Management (DPSM) automation that can be extended in several promising directions:

\textbf{Cloud AI Services Integration}: Future implementations can leverage cloud-based AI services such as AWS Bedrock for foundation model access, Amazon SageMaker for custom model training and deployment, and Azure OpenAI Service for enhanced natural language processing capabilities.

\textbf{Enhanced Agent Toolsets}: The current framework can be expanded with additional tools and capabilities including database search agents for querying operational data repositories, web search agents for accessing real-time documentation and knowledge bases, and specialized agents for log analysis, performance metrics correlation, and security event processing.

\textbf{Hierarchical Agent Architecture}: A multi-tier agent architecture utilizing both A2A and MCP protocols could enable more sophisticated coordination patterns, including supervisor agents that coordinate multiple specialized agents and domain-specific agent clusters that handle complex multi-step workflows across different ITIL practices.

\textbf{Retrieval Augmented Generation (RAG) Implementation}: Integration of RAG capabilities would enable agents to access and reason over large knowledge bases including ITIL documentation, organizational runbooks, historical incident data, and best practice repositories.

\section{Conclusion}

This implementation demonstrates the practical application of AIOps principles in automating ITIL v5 service management practices through intelligent agent coordination. The A2A protocol enables standardized communication between specialized agents, facilitating automated incident management, infrastructure monitoring, and service request fulfillment. The framework successfully addresses key challenges in traditional monitoring and event management by providing autonomous event correlation, intelligent ticket routing, and seamless integration between Kubernetes infrastructure monitoring and ITIL-compliant service desk operations.

The research contributes to the field of Agentic AIOps by demonstrating a working implementation that combines multiple cutting-edge technologies including LangGraph for agent orchestration, MCP for standardized tool integration, and A2A protocols for inter-agent communication. The distributed architecture presented in this work provides a scalable foundation for enterprise IT operations automation while maintaining the flexibility to integrate with existing ITSM tools and processes.

\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{acharya2025}
Acharya, D. B., Kuppan, K., \& Divya, B. (2025). Agentic AI: Autonomous intelligence for complex goals. \textit{IEEE}.

\bibitem{anthropic2024}
Anthropic. (2024). \textit{Model Context Protocol (MCP)}. Retrieved from \url{https://modelcontextprotocol.io/}

\bibitem{chen2024}
Chen, L., Wang, X., \& Liu, Y. (2024). Multi-agent systems for cloud infrastructure management: A comprehensive framework. \textit{Journal of Cloud Computing}, 13(2), 45-62.

\bibitem{johnson2024}
Johnson, M., Davis, P., \& Anderson, R. (2024). Human-AI collaboration patterns in DevOps: Factors for successful integration. \textit{IEEE Software}, 41(3), 78-87.

\bibitem{kumar2023}
Kumar, S., \& Singh, R. (2023). Deep learning approaches for anomaly detection in distributed systems. \textit{IEEE Transactions on Network and Service Management}, 20(4), 1823-1836.

\bibitem{langchain2024}
LangChain. (2024). \textit{LangGraph: Build language agents as graphs}. Retrieved from \url{https://langchain-ai.github.io/langgraph/}

\bibitem{li2023}
Li, Q., \& Zhang, W. (2023). AI-powered monitoring systems: Combining machine learning with rule-based approaches for intelligent event management. \textit{Journal of Network and Computer Applications}, 201, 103-118.

\bibitem{openai2024}
OpenAI. (2024). \textit{Agent2Agent Protocol Specification}. Retrieved from \url{https://github.com/openai/agent2agent}

\bibitem{piccialli2025}
Piccialli, F., Chiaro, D., Sarwar, S., Cerciello, D., Qi, P., \& Mele, V. (2025). AgentAI: A comprehensive survey on autonomous agents in distributed AI for industry 4.0.

\bibitem{rodriguez2023}
Rodriguez, A., \& Martinez, C. (2023). Intelligent ticketing systems using natural language processing for ITIL automation. \textit{International Journal of Information Management}, 71, 102-115.

\bibitem{thompson2024}
Thompson, K., Brown, D., \& Wilson, S. (2024). Secure agent-to-agent communication protocols for distributed computing environments. \textit{Computer Networks}, 228, 109-123.

\bibitem{wang2024}
Wang, H., Zhang, M., \& Li, J. (2024). Reinforcement learning for automated incident response in IT operations. \textit{ACM Transactions on Autonomous and Adaptive Systems}, 18(1), 1-24.

\end{thebibliography}

\end{document}
